---
title: "ML Project"
author: "Stephen Trask"
date: "Saturday, July 26, 2014"
output: word_document
---

##Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Objective
The goal is to predict the manner in which respondents did the exercise. This is the "classe" variable in the training set.

##Data cleaning
There are two dataset to achieve the objective, 
.       training data (pml-training.csv) 
.        testing data (pml-testing.csv)
In the raw data there are several variables with lots of missing (NA) values. Our first task in data cleaning is to calculate percentage of NA values in each column of the dataset and then keep only those variables with less than 10% of NA values. By performing this step we will have a dataset with all variables with complete records. The next step is the check the outcome variable of interest, whether there is any unexpected class level present or not, if there is any then we will drop those rows from the dataset. We also dropped those categorical variables which have very sparse observation across classes. After performing all of these operations we will have a cleaned data in hand.

```{r}
# Opening training and test data
trdata <- read.csv("pml-training.csv",stringsAsFactors=FALSE,na.strings = "NA")
testdata <- read.csv("pml-testing.csv",stringsAsFactors=FALSE,na.strings = "NA")

# Combine both training and test data to perform pre-processing at once
# The following pre-processing will be performed
# 1) Drop variables from analysis if there is more than xxx percentage of NA value present
# 2) If there are any NA value presents then we will impute with median for numeric variable and mode for categorical variables

# Dropping problem_id variable from test data
testdata <- testdata[,-which(names(testdata)=="problem_id")]

# Adding the classe variable in test data with class X, we will remove this after pre processing step
testdata$classe<-"X"

# Adding new variable to track training and test data
trdata$traintest <- 1
testdata$traintest <-0

# Combining both dataset
alldata <- rbind(trdata,testdata,stringsAsFactors=F)

# Calculating percentage of NA values in a variable
napct <- apply(alldata,2,function(x) 100*sum(is.na(x))/length(x))
blankpct <- apply(alldata,2,function(x) 100*sum(nchar(x)==0)/length(x))

# Index of the variable that will be dropped due to excess NA values
dropvarIndex <- c(which(napct>10),which(blankpct>10))
length(dropvarIndex)

# dropping the variables with more than 10% NA values
alldata <- alldata[,-dropvarIndex]

# dropping categorical variables with very sparse data
table(alldata$new_window)
alldata <- alldata[,-which(names(alldata)=="new_window")]

# Dropping the variable with name X because this is just the serial number from the csv file
alldata <- alldata[,-which(names(alldata)=="X")]

# Checking distribution of classe variable
table(alldata$classe)

# Dropping observation with classe==FALSE
alldata <- alldata[-which(alldata$classe=="FALSE"),]

# Now separating out the training and test data

trainingData <- alldata[alldata$traintest==1,]
trainingData <- trainingData[,-which(names(trainingData)=="traintest")]

testData <- alldata[alldata$traintest==0,]
testData <- testData[,-which(names(testData)=="traintest")]
testData <- testData[,-which(names(testData)=="classe")]

# extracting only the variable names that will be used in modeling stage
modelingVar <- names(trainingData)[-c(1:4)]

# converting the outcome variable into factor mode for analysis
trainingData$classe <- as.factor(trainingData$classe)
```

##Model building
Since we have only the training and test data, we need another dataset to see the model performance. To do so, we will split the initial training data into two parts with 75% in the final training data and 25% will be in validation data. With the 75% of the training data we will build a series of K-nearest neighbor classification model with k varies from 5 to 13 with increment 2. We then select best model with highest accuracy with k=5. We have used 10-fold cross validation to estimate the cross validation error. Finally, we also estimated the accuracy in the validation set and then perform the prediction on test data.


```{r}

# Creating training and validation data by splitting initial training data
library(caret)
set.seed(4929)
inTrain <- createDataPartition(trainingData$classe, p = 3/4)[[1]]
training <- trainingData[ inTrain,]
validation <- trainingData[-inTrain,]

# Model building

# Setting model control parameter with 10-fold cross validation
ctrl <- trainControl(method = "cv",number=10)

# Fitting the knn model with k=5, though we have used different k values (5,7,9,11,13) but in we kept only the best one with tuneLength=1
model_1 <- train(classe~.,data=training[modelingVar],method="knn",tuneLength = 1,trControl = ctrl)

```

```{r}
# printing model output
model_1

```

### Results
The estimated 10-fold cross validation error is 8.5%. The accuracy statistics based on the validation is:
```{r}
predict_1 <- predict(model_1,newdata=validation[modelingVar[-length(modelingVar)]])
```

```{r}
### Calculating accuracy statistics based on the prediction on validation data
confusionMatrix(data=predict_1,reference=validation$classe)
```

```{r}
### Prediction on test data
predict_test <- predict(model_1,newdata=testData[modelingVar[-length(modelingVar)]])
```

```{r}
### Preparing test data with predicted class levels
predicted_test_data <- data.frame(testData,classe=predict_test)
write.csv(predicted_test_data,file="predicted_test_data.csv",row.names=F)
```











